Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-08-22 paper.qmd
MSE Vs Counts

  ## Compute error
  error_df <- glance(
    df_bin_centroids = df_bin_centroids_scurve,
    df_bin = df_bin_scurve,
    training_data = training_data_scurve,
    newdata = NULL,
    type_NLDR = "UMAP",
    col_start = "x") |>
    mutate(bin1 = xbins,
           bin2 = bin2,
           b = bin1 * bin2,
           b_non_empty = NROW(df_bin_centroids_scurve),
           a1 = a1)

@@

a1_m_scurve <- ggplot(error_scurve,
                     aes(x = a1,
                         y = b_non_empty)) +
  geom_point(size = 1) +
  geom_line(linewidth = 0.05) +
  geom_vline(xintercept = 0.116,
           linetype=2,
           color = "#bdbdbd",
           linewidth=1) +
  geom_hline(yintercept = 52,
           linetype=2,
           color = "#bdbdbd",
           linewidth=1) +
  # geom_vline(xintercept = 11, linetype="solid",
  #            color = "#bdbdbd", linewidth=1) +
  # geom_vline(xintercept = 9, linetype=2,
  #            color = "#bdbdbd", linewidth=1) +
  # geom_vline(xintercept = 13, linetype=2,
  #            color = "#bdbdbd", linewidth=1) +
  scale_y_continuous(breaks = sort(unique(error_scurve$b_non_empty))) +
  scale_x_continuous(breaks = sort(unique(round(error_scurve$a1, 3)))) +
  labs(x = expression(a[1]), y = expression(m)) +
  ggtitle("(b)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 10,
                                   angle = 90),
        axis.text.y = element_text(size = 11),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12,
            angle=90),
        axis.ticks.x = element_line(linewidth = 0.5),
        axis.ticks.y = element_line(linewidth = 0.5),
        plot.title = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10),
        legend.key.height = unit(1, 'cm'),
        legend.key.width = unit(1, 'cm'),
        legend.key.size = unit(1, 'cm'))
# 
# b1_m_scurve2 <- ggplot(error_scurve, 
#                      aes(x = bin1, 
#                          y = b_non_empty)) + 
#   geom_point(size = 1) +
#   geom_line(linewidth = 0.05) + 
#   # geom_vline(xintercept = 11, linetype="solid",
#   #            color = "#bdbdbd", linewidth=1) +
#   # geom_vline(xintercept = 9, linetype=2,
#   #            color = "#bdbdbd", linewidth=1) +
#   # geom_vline(xintercept = 13, linetype=2,
#   #            color = "#bdbdbd", linewidth=1) +
#   scale_x_continuous(breaks = 5:19) +
#   labs(x = expression(b[1]), y = expression(m)) +
#   ggtitle("(c)") +
#   theme(aspect.ratio = 0.75,
#         axis.text.x = element_text(size = 12),
#         axis.text.y = element_text(size = 12),
#         axis.title.x = element_text(size = 12),
#         axis.title.y = element_text(size = 12, 
#             angle=90),
#         axis.ticks.x = element_line(linewidth = 0.5),
#         axis.ticks.y = element_line(linewidth = 0.5),
#         plot.title = element_text(size = 12),
#         legend.title = element_text(size = 12),
#         legend.text = element_text(size = 10),
#         legend.key.height = unit(1, 'cm'),
#         legend.key.width = unit(1, 'cm'),
#         legend.key.size = unit(1, 'cm'))
```

@@

<!--distribution of distance along with the default benchmark-->

```{r}
#| label: diff-bins-scurve-buff3
#| eval: false
## To initialize benchmark values to remove low density hexagons
benchmark_rm_hex_vec <- seq(0, 0.7, by=0.1)
bin1_vec_scurve <- c(7, 11, 17)
distance_all_scurve <- data.frame(matrix(nrow = 0, ncol = 0))
#buffer_vec <- 0.07

for (xbins in bin1_vec_scurve) {
  for (buff in buffer_vec) {

  #for (benchmark_rm_lwd in benchmark_rm_hex_vec) {

    a1 <- calc_bins_y(bin1 = xbins, r2 = r2, q = buff)$a1

    scurve_model <- fit_highd_model(
      training_data = training_data_scurve,
      emb_df = umap_scurve_scaled,
      bin1 = xbins,
      r2 = r2,
      q = buff, 
      is_bin_centroid = TRUE,
      is_rm_lwd_hex = FALSE,
      col_start_highd = "x"
    )

    tr1_object_scurve <- tri_bin_centroids(
      df_bin_centroids_scurve_high_dens, 
      x = "c_x", 
      y = "c_y")

    tr_from_to_df_scurve <- gen_edges(
      tri_object = tr1_object_scurve)

    start_point <- a1/2

    # Create bins and calculate the mean value for each bin range
    bins <- seq(start_point, max(distance_scurve$distance) + a1, 
                by = a1)
    bin_labels <- bins[-length(bins)] + a1 / 2  # mean of each bin range

    ## Compute error
    distance_scurve <- distance_scurve |>
      mutate(q = buff,
             bin1 = xbins,
             a1 = a1)

    distance_all_scurve <- bind_rows(distance_all_scurve, distance_scurve)

# dist_summary_scurve <- distance_all_scurve |>
#     count(bin1, q, bin)

distance_hist <- distance_all_scurve |>
  filter(bin1 == 11) |>
  filter(q == 0.07) |>
  ggplot(aes(x = distance)) +
    geom_histogram(
      aes(y = after_stat(count) / sum(after_stat(count)), 
      group = as.factor(q),
      fill = as.factor(q)),
      binwidth = a1,
      position = "dodge") +
  scale_x_continuous(breaks = bin_labels, labels = round(bin_labels, 3)) +
  scale_color_brewer(type = "qual") +
  #facet_wrap(vars(as.factor(bin1)), scale = "free") +
  ylab("standardized number of edges") +
  xlab(expression(d^{(2)})) +
  ggtitle("(d)") +
  theme(aspect.ratio = 0.75,
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12, angle = 90),
        axis.text.x = element_text(size = 12, angle = 90),
        axis.text.y = element_text(size = 12),
        axis.ticks.x = element_line(), 
        plot.title = element_text(size = 12))

```{r}
#| label: triangulate-scurve2

@@

distance_hist <- ggplot(
  distance_scurve, aes(x = distance)) +
    geom_histogram(
      aes(y = after_stat(count) / sum(after_stat(count))), 
      binwidth = bin_width) +
  scale_x_continuous(breaks = bin_labels, labels = round(bin_labels, 3)) +
  ylab("standardized number of edges") +
  xlab(expression(d^{(2)})) +
  ggtitle("(d)") +
  theme(aspect.ratio = 0.75,
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12, angle = 90),
        axis.text.x = element_text(size = 12, angle = 90),
        axis.text.y = element_text(size = 12),
        axis.ticks.x = element_line(), 
        plot.title = element_text(size = 12))
```

@@

mse_scurve_b + a1_m_scurve +
  mse_scurve_lwd + distance_hist + 
  plot_layout(ncol=2)
```

>>>>

  ## Compute error
  error_df <- glance(
    df_bin_centroids = df_bin_centroids_scurve,
    df_bin = df_bin_scurve,
    training_data = training_data_scurve,
    newdata = NULL,
    type_NLDR = "UMAP",
    col_start = "x") |>
    mutate(bin1 = xbins,
           bin2 = bin2,
           b = bin1 * bin2,
           b_non_empty = NROW(df_bin_centroids_scurve),
           a1 = a1,
           mean_counts = sum(df_bin_centroids_scurve$bin_counts)/NROW(df_bin_centroids_scurve))

@@

a1_m_scurve <- ggplot(error_scurve,
                     aes(x = a1,
                         y = b_non_empty)) +
  geom_point(size = 1) +
  geom_line(linewidth = 0.05) +
  geom_vline(xintercept = 0.116,
           linetype=2,
           color = "#bdbdbd",
           linewidth=1) +
  geom_hline(yintercept = 52,
           linetype=2,
           color = "#bdbdbd",
           linewidth=1) +
  # geom_vline(xintercept = 11, linetype="solid",
  #            color = "#bdbdbd", linewidth=1) +
  # geom_vline(xintercept = 9, linetype=2,
  #            color = "#bdbdbd", linewidth=1) +
  # geom_vline(xintercept = 13, linetype=2,
  #            color = "#bdbdbd", linewidth=1) +
  scale_y_continuous(breaks = sort(unique(error_scurve$b_non_empty))) +
  scale_x_continuous(breaks = sort(unique(round(error_scurve$a1, 3)))) +
  labs(x = expression(a[1]), y = expression(m)) +
  ggtitle("(b)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 10,
                                   angle = 90),
        axis.text.y = element_text(size = 11),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12,
            angle=90),
        axis.ticks.x = element_line(linewidth = 0.5),
        axis.ticks.y = element_line(linewidth = 0.5),
        plot.title = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10),
        legend.key.height = unit(1, 'cm'),
        legend.key.width = unit(1, 'cm'),
        legend.key.size = unit(1, 'cm'))

count_mse_scurve <- ggplot(error_scurve,
                     aes(x = mean_counts,
                         y = MSE)) +
  geom_point(size = 1) +
  geom_line(linewidth = 0.05) +
  # geom_vline(xintercept = 0.116,
  #          linetype=2,
  #          color = "#bdbdbd",
  #          linewidth=1) +
  # geom_hline(yintercept = 52,
  #          linetype=2,
  #          color = "#bdbdbd",
  #          linewidth=1) +
  # geom_vline(xintercept = 11, linetype="solid",
  #            color = "#bdbdbd", linewidth=1) +
  # geom_vline(xintercept = 9, linetype=2,
  #            color = "#bdbdbd", linewidth=1) +
  # geom_vline(xintercept = 13, linetype=2,
  #            color = "#bdbdbd", linewidth=1) +
  labs(x = expression(mean(n[h])), y = "MSE") +
  ggtitle("(c)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 10,
                                   angle = 90),
        axis.text.y = element_text(size = 11),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12,
            angle=90),
        axis.ticks.x = element_line(linewidth = 0.5),
        axis.ticks.y = element_line(linewidth = 0.5),
        plot.title = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10),
        legend.key.height = unit(1, 'cm'),
        legend.key.width = unit(1, 'cm'),
        legend.key.size = unit(1, 'cm'))

# 
# b1_m_scurve2 <- ggplot(error_scurve, 
#                      aes(x = bin1, 
#                          y = b_non_empty)) + 
#   geom_point(size = 1) +
#   geom_line(linewidth = 0.05) + 
#   # geom_vline(xintercept = 11, linetype="solid",
#   #            color = "#bdbdbd", linewidth=1) +
#   # geom_vline(xintercept = 9, linetype=2,
#   #            color = "#bdbdbd", linewidth=1) +
#   # geom_vline(xintercept = 13, linetype=2,
#   #            color = "#bdbdbd", linewidth=1) +
#   scale_x_continuous(breaks = 5:19) +
#   labs(x = expression(b[1]), y = expression(m)) +
#   ggtitle("(c)") +
#   theme(aspect.ratio = 0.75,
#         axis.text.x = element_text(size = 12),
#         axis.text.y = element_text(size = 12),
#         axis.title.x = element_text(size = 12),
#         axis.title.y = element_text(size = 12, 
#             angle=90),
#         axis.ticks.x = element_line(linewidth = 0.5),
#         axis.ticks.y = element_line(linewidth = 0.5),
#         plot.title = element_text(size = 12),
#         legend.title = element_text(size = 12),
#         legend.text = element_text(size = 10),
#         legend.key.height = unit(1, 'cm'),
#         legend.key.width = unit(1, 'cm'),
#         legend.key.size = unit(1, 'cm'))
```

@@

<!--distribution of distance along with the default benchmark-->
```{r}
#| label: triangulate-scurve2

@@

distance_hist <- ggplot(
  distance_scurve, aes(x = distance)) +
    geom_histogram(
      aes(y = after_stat(count) / sum(after_stat(count))), 
      binwidth = bin_width) +
  scale_x_continuous(breaks = bin_labels, labels = round(bin_labels, 3)) +
  ylab("standardized number of edges") +
  xlab(expression(d^{(2)})) +
  ggtitle("(d)") +
  theme(aspect.ratio = 0.75,
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12, angle = 90),
        axis.text.x = element_text(size = 12, angle = 90),
        axis.text.y = element_text(size = 12),
        axis.ticks.x = element_line(), 
        plot.title = element_text(size = 12))

distance_scurve_summary <- distance_scurve |>
  group_by(bin) |>
  summarise(count = n()) |>
  mutate(std_counts = count/sum(count))

@@

mse_scurve_b + a1_m_scurve +
  count_mse_scurve + distance_hist + 
  plot_layout(ncol=2)
```
